---
title: "NorwaySDMs: R package to produce species distribution models in a reporducible framework"
output: html_notebook
---

In support of our manuscript, we developed an R package to help construct point-process species distribution models (SDMs) from disperate in a simple and reproducible framework. This *RMarkdown* document presents an illustration of the package by creating an SDM for red-listed plant species obtained via the *Vascular Plant Field Notes* in Norway, as well as citizen-science data obtained from *Global Biodiversity Information Facility* (*GBIF*). The first step in exploring this document is to download the package using the following script:

```{r Install package, warning = FALSE, message = FALSE}

devtools::install_github('PhilipMostert/NorwaySDMs')
library(NorwaySDMs)

```

As well as any of the following packages required:

```{r Load packages, warning = FALSE, message = FALSE}

library(ggplot2)
library(sf)
library(sp)
library(plyr)
library(dplyr) 
library(ggmap) 
library(maps)
library(PointedSDMs)
library(inlabruSDMs)
library(spatstat)
library(maptools)
library(INLA)
library(rgeos)
library(fields)
library(viridis)
library(RColorBrewer)

```

And finally by loading in some objects which will be required later on.

```{r Norway map, warning = FALSE, message = FALSE}

Projection <- CRS('+proj=utm +zone=32 +ellps=WGS84 +datum=WGS84 +units=m +no_defs')

norwayfill <- map("world", "norway", fill=TRUE, plot=FALSE, 
                  ylim=c(58,72), xlim=c(4,32))
IDs <- sapply(strsplit(norwayfill$names, ":"), function(x) x[1])
norway.poly <- map2SpatialPolygons(norwayfill, IDs = IDs, 
                                   proj4string = Projection)

```

Before conducting any analysis, we explore and make some adjustments to the data which we will use later in our integrated model. The *Vascular Plant Field Notes* is ... (ADD LATER).

```{r Read PA data, warning = FALSE, message = FALSE}

norge_data <- read.csv('./data/VU.PA.data.Norge.txt', sep = ' ')

olso_data <- read.csv('./data/VU.PA.data.Olso.txt', sep = ' ')
olso_data <- olso_data[!(olso_data$decimalLongitude == 0),] 
oslo_data <- olso_data[!(olso_data$decimalLatitude == 0),]

PA_data <- rbind(norge_data, olso_data)

```

In this analysis, we chose the three most abundant species available in our dataset. This was done mainly to make the example more computationally friendly, but also because a significant proportion of the plants in the dataset only had a handful of observations (a possible consequence of working with red-listed data), making inference on those species difficult.

```{r Select three species, warning = FALSE, message = FALSE}

abundant <- PA_data %>% group_by(scientificName) %>%
                        count() %>% 
                        arrange(desc(freq)) %>%
                        data.frame
abundant <- abundant[1:3, 'scientificName']

PA_data <- PA_data[PA_data$scientificName%in%abundant,]

PA_data <- sp::SpatialPointsDataFrame(coords = data.frame(PA_data$decimalLongitude,                                                PA_data$decimalLatitude),
                     data = data.frame(scientificName = PA_data$scientificName, 
                                      individualCount = PA_data$individualCount),
                                      proj4string = Projection)

```

```{r Plot of PA, warning = FALSE, message = FALSE}

ggplot() + 
  gg(norway.poly) + 
  gg(PA_data) # + others...


```

We treated the *Vascular Plant Field Notes* as detection/non-detection data by creating a gridded map of sampling locations across our Norway polygon and noting which sampling locations contained which species of plants. That is, we treated these data as a Bernoulli random variable taking on unity if the species was detected in a given grid, otherwise taking on zero if it was not detected. To do this, we used a nearest neighbour classifier to assign each of the plant species' location to its nearest grid point.

```{r Plot of grid, warning = FALSE, message = FALSE}

grid <- makegrid(norway.poly, cellsize = 0.25, pretty = FALSE)
grid <- SpatialPoints(grid, proj4string = Projection)
ggplot() +
  gg(grid) + 
  gg(PA_data) + 
  gg(norway.poly) #+ others...

spgrdWithin <- SpatialPixels(grid[norway.poly,])
spgrdWithin <- as(spgrdWithin, "SpatialPolygons")
ggplot() + 
  gg(spgrdWithin) #+ others...

```

```{r Assign species to nearest grid, warning = FALSE, message = FALSE}

PA_data_frame <- as.data.frame(PA_data)
closest <- RANN::nn2(data = grid@coords, query = PA_data_frame[,3:4], k = 1)
grid$ID <- seq(nrow(grid@coords))

closest_ind <- as.data.frame(closest) %>%
  dplyr::rename(ID = nn.idx)

closest_ind$scientificName <- PA_data_frame$scientificName
closest_ind$individualCount <- PA_data_frame$individualCount

joined <- dplyr::inner_join(closest_ind,as.data.frame(grid), by = 'ID') 

grid_data <- sp::SpatialPointsDataFrame(coords = data.frame(joined$x1, joined$x2),
                                        data = data.frame(scientificName = joined$scientificName, individualCount = joined$individualCount),
                                        proj4string = Projection)
colnames(grid_data@coords) <- c('x','y')

```

```{r Plot of gridded data, warning = FALSE, message = FALSE}

grid_data <- grid_data[!is.na(over(grid_data, norway.poly)),]
ggplot() + gg(norway.poly) + gg(grid_data, col = 'red') #+ gg(grid)

```

```{r Create grid index, warning = FALSE, message = FALSE}

species <- unique(grid_data$scientificName)

unique_grid <- as.data.frame(grid_data) %>%
  group_by(x,y) %>%
  slice(1) %>%
  dplyr::select(x,y) %>%
  data.frame()

grid_index <- list()
for (i in 1:nrow(unique_grid)) {
  
  x <- unique_grid[i,1]
  y <- unique_grid[i,2]
  
  present <- grid_data[grid_data@coords[,1] == x & grid_data@coords[,2] == y,]
  
  not_in <- species[!species%in%present$scientificName]
  
  if (!identical(not_in, character(0))) {
    
    ind <- data.frame(scientificName = not_in, individualCount = 0)
    
    ind_coords <- data.frame(x,y) %>% slice(rep(1:n(), each = length(not_in)))
    
    absent <- sp::SpatialPointsDataFrame(coords = ind_coords, data = ind,
                                         proj4string = Projection)
    
    grid_index[[i]] <- rbind.SpatialPointsDataFrame(present, absent)
    
  }
  else grid_index[[i]] <- present
  
}

```

```{r Plot of PA data in grid, warning = FALSE, message = FALSE}


PA_data <- do.call(rbind.SpatialPointsDataFrame, grid_index)

ggplot() + gg(PA_data, aes(col = factor(individualCount))) +
  facet_grid(~scientificName) +
  gg(norway.poly) +
  gg(spgrdWithin) +
  coord_equal() +
  scale_fill_continuous(guide = guide_legend()) +
  scale_color_manual(labels = c('Absent', "Present"), values = c("#d11141", "#00aedb")) +
  labs(x = 'Longitude', y = 'Latitude', col = 'Grid Observation') +
  ggtitle('Present absence data') +
  theme_classic() +
  theme(legend.position="bottom",
        plot.title = element_text(hjust = 0.5))

```

```{r Structured data, warning = FALSE, message = FALSE}

colnames(PA_data@coords) <- c('longitude','latitude')
PA_data$scientificName <- gsub(' ','_',PA_data$scientificName)
names(PA_data@data) <- c('species', 'individualCount')

structured <- structured_data(PA_data, datasetType = 'PA',
                             responsePA = 'individualCount',
                             coordinateNames = colnames(PA_data@coords))

```

```{r Mesh construction, warning = FALSE, message = FALSE}

mesh <- species_model(speciesNames = unique(structured@dataPA$PA_data$species),
                               structuredData = structured,
                               boundary = norway.poly,
                               return = 'mesh', limit = 5000,
meshParameters = list(cutoff=0.08, max.edge=c(1, 3), offset=c(1,1))

ggplot() + gg(mesh)

```

```{r All species plot, warning = FALSE, message = FALSE}

species_plot <- species_model(speciesNames = unique(structured@dataPA$PA_data$species),
                               structuredData = structured,
                               boundary = norway.poly,
                               return = 'species plot', limit = 5000, mesh = mesh)
ggplot() + gg(species_plot)

```

```{r Prediction maps, warning = FALSE, message = FALSE}

prediction_maps <- species_model(speciesNames = unique(structured@dataPA$PA_data$species),                                    scale = TRUE, structuredData = structured,
                               worldclimCovariates = 'Annual Mean Temperature', 
                               boundary = norway.poly,
                               return = 'predictions map', limit = 5000,
                               mesh = mesh)

ggplot() + gg(prediction_maps)

```
