---
title: "Building a worklfow for Warblers acorss Pennsylvania state"
output: rmarkdown::html_vignette
bibliography: '`r system.file("References.bib", package="intSDM")`'
vignette: >
  %\VignetteIndexEntry{PennsylvaniaWarbler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  fig.width=8, 
  fig.height=5,
  warning = FALSE,
  message = FALSE)
```

This vignette illustrates using the *intSDM* R package for three types of warbler distributed across Pennsylvania on the Eastern side of the United States of America. This case study has been used in numerous other integrated species distribution model analyses and includes three datasets: *eBird*, *North American Breeding Bird Survey* (*BBS*) and *Pennsylvania Breeding Bird Atlas* (*BBA*). Details on the data and the selection of observation models for each are provided in @PointedSDMs, @isaac2020data and @miller2019recent.

```{r setup}

library(intSDM)
library(USAboundaries)

```

We will assume that the *BBA* and *BBS* data are not provided on *GBIF*, and thus will load them directly from the *PointedSDMs* package.

```{r get data}

BBA <- PointedSDMs::SetophagaData$BBA
BBA$Species_name <- paste0('Setophaga_', BBA$Species_name)
BBS <- PointedSDMs::SetophagaData$BBS
BBS$Species_name <- paste0('Setophaga_', BBS$Species_name)

```

We will then initialize the workflow using the `startWorkflow` function.

```{r startWorkflow}

workflow <- startWorkflow(
  Projection = "WGS84",
  Species = c("Setophaga_caerulescens", "Setophaga_fusca", "Setophaga_magnolia"),
  saveOptions = list(projectName =  'Setophaga'), Save = FALSE
)

```

The `.$addArea()` function only gives us access to country borders. However we can easily add other polygon objects to the workflow using the *Object* argument from the function.

```{r addArea}

workflow$addArea(Object = USAboundaries::us_states(states = "Pennsylvania"))

```

Next we add data to the analysis. The *eBird* dataset is available to download directly from *GBIF*, and thus may be downloaded into out workflow using the `.$addGBIF` function and specifying the relevant *datasetKey*. The other two datasets are not directly available on *GBIF*, but may still be added using the `.$addStructured` function. This requires us to specify the response name of each dataset (using the *responseName* argument), and the species name variable (using the *speciesName* argument).

```{r download Data}

workflow$addGBIF(datasetName = 'eBird', datasetType = 'PO', limit = 5000,
                 datasetKey = '4fa7b334-ce0d-4e88-aaae-2e0c138d049e')

workflow$addStructured(dataStructured = BBA, datasetType = 'PA',
                       responseName = 'NPres', 
                       speciesName = 'Species_name')

workflow$addStructured(dataStructured = BBS, datasetType = 'Counts',
                       responseName = 'Counts', 
                       speciesName = 'Species_name')

workflow$plot(Species = TRUE)

```

We can then add the *elevation* and *canopy* covariates from the *PointedSDMs* package. If we were planning on using *worldClim* covariates, we could have used the *worldClim* argument by specifying which variable we wanted to download.

```{r addCovariates}

covariates <- scale(terra::rast(system.file('extdata/SetophagaCovariates.tif', 
                                      package = "PointedSDMs")))
names(covariates) <- c('elevation', 'canopy')

workflow$addCovariates(Object = covariates)
  
workflow$plot(Covariates = TRUE)

```

We specify an additional field for the *eBird* dataset using the `.$biasFields` argument, and an *inla.mesh* object using `.$addMesh`.

```{r biasFields}

workflow$biasFields(datasetName  = 'eBird')

workflow$addMesh(cutoff = 0.2,
                 max.edge = c(0.1, 0.24),
                 offset = c(0.1, 0.4))

```

For this case study, we specify the model outcome as *Model*. This will give us the *R-INLA* model outcome for which we could analyse further.

```{r outcomes}

workflow$workflowOutput('Model')

workflow$modelOptions(INLA = list(control.inla=list(int.strategy = 'eb',
                                                    cmin = 0),
                                  safe = TRUE,
                                  inla.mode = 'experimental'))

```

```{r sdmWorkflow}

Models <- sdmWorkflow(Workflow = workflow)

lapply(unlist(Models, recursive = FALSE), summary)

```
